---
title: Unveiling the impact of bacteriophages targeting Shiga toxin-encoding Escherichia
  coli on the colon microbiome in vitro fermentation
author: "Sudarshan and Graça Pinto"
date: "`r format(Sys.Date(), "%B %d, %Y")`"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
    code_folding: hide
subtitle: PFU, CFU and fermentation analysis
editor_options:
  chunk_output_type: console
---


# Input data into R studio
```{r}
library(data.table)
library(purrr)
library(dplyr)
library(reshape2)
library(microbiomeutilities)
library(ggplot2)
library(ggpubr)
mycols <- c(Control= "brown3", 
            E.coli= "steelblue",
            Phage = "grey50", 
            Phage_E.coli = "#7c65a9")
mycolsgit <- c(Control="grey50",
               GIT="steelblue")
```

## read metadata

```{r}
meta_tabl <- read.table(file = "data/metadata_v1.txt", 
                            header = T, 
                            quote = "", 
                            sep = "\t", stringsAsFactors = F)
rownames(meta_tabl) <- meta_tabl$AssayIdentifier

study <- c("D1", "D1", "D1", "D1", "D1", "D1", "D1", "D1",
           "D2", "D2", "D2", "D2", "D2", "D2", "D2", "D2",
           "D3", "D3", "D3", "D3", "D3", "D3", "D3", "D3")
meta_tabl$study <- study
```


## PFU Dynamics    
```{r}
PFU.inv <- as.data.frame(read.csv("data/raw/wetlab/PFUinv.csv", sep = ";"))
colnames(PFU.inv)[1] <- "AssayIdentifier"

PFU.inv.mlt <- reshape2::melt(PFU.inv) %>% 
  mutate(Time = as.numeric(gsub("X", "", variable))) %>% 
  select(-variable) %>%
  left_join(meta_tabl)
PFU.inv.mlt$Perturbation <-  gsub( "Perturbation ", "",PFU.inv.mlt$Condition)

```

### Plot PFU  
```{r fig.width=6, fig.height=4}
PFUplot <- ggplot(PFU.inv.mlt, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylim(0,8) +
  ylab("Log10 PFU per mL")
ggsave("01_wetlab/figs/pfu_dynamics.pdf", h=4, w=8)
ggsave("01_wetlab/figs/pfu_dynamics.png", h=4, w=8, dpi = 300)

ggplot(PFU.inv.mlt, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylab("Log10 PFU per mL") +
  facet_grid(~study, scales = "free")
ggsave("01_wetlab/figs/pfu_dynamics_study.pdf", h=4, w=8)
```

## CFU dynamics  

```{r}
CFU.inv <- as.data.frame(read.csv("data/raw/wetlab/CFUinv.csv", sep = ";"))
colnames(CFU.inv)[1] <- "AssayIdentifier"

CFU.inv.mlt <- reshape2::melt(CFU.inv) %>% 
  mutate(Time = as.numeric(gsub("t", "", variable))) %>% 
  select(-variable) %>% 
  left_join(meta_tabl)
CFU.inv.mlt$Perturbation <-  gsub( "Perturbation ", "",CFU.inv.mlt$Condition)

#CFU.inv.mlt$Time
```

### Plot CFU  
```{r fig.width=6, fig.height=4}
CFUplot <- ggplot(CFU.inv.mlt, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylim(0,8) +
  ylab("Log10 CFU per mL")
ggsave("01_wetlab/figs/cfu_dynamics.pdf", h=4, w=8)
ggsave("01_wetlab/figs/cfu_dynamics.png", h=4, w=8, dpi = 300)

ggplot(CFU.inv.mlt, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylab("Log10 CFU per mL") +
  facet_grid(~study, scales = "free")
ggsave("01_wetlab/figs/cfu_dynamics_study.pdf", h=4, w=8)
```

## CO2 dynamics  

```{r}
GC.CO2.inv <- as.data.frame(read.csv("data/raw/wetlab/GCCO2inv.csv", sep = ";"))
colnames(GC.CO2.inv)[1] <- "AssayIdentifier"

GC.CO2.inv.mlt <- reshape2::melt(GC.CO2.inv) %>% 
  mutate(Time = as.numeric(gsub("t", "", variable))) %>% 
  select(-variable) %>% 
  left_join(meta_tabl)
GC.CO2.inv.mlt$Perturbation <-  gsub( "Perturbation ", "",GC.CO2.inv.mlt$Condition)

#CFU.inv.mlt$Time
```

### Plot CO2  
```{r fig.width=6, fig.height=4}
ggplot(GC.CO2.inv.mlt, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylab("% CO2")
ggsave("01_wetlab/figs/co2_dynamics.pdf", h=4, w=8)

co2 <- ggplot(GC.CO2.inv.mlt, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylab("% CO2") +
  facet_grid(~study, scales = "free")
ggsave("01_wetlab/figs/co2_dynamics_study.pdf", h=4, w=8)
ggsave("01_wetlab/figs/co2_dynamics_study.png", h=4, w=8, dpi = 300)

```

## H2 dynamics  

```{r}
GC.H2.inv <- as.data.frame(read.csv("data/raw/wetlab/GCH2inv.csv", sep = ";"))
colnames(GC.H2.inv)[1] <- "AssayIdentifier"

GC.H2.inv.mlt <- reshape2::melt(GC.H2.inv) %>% 
  mutate(Time = as.numeric(gsub("t", "", variable))) %>% 
  select(-variable) %>% 
  left_join(meta_tabl)
GC.H2.inv.mlt$Perturbation <-  gsub( "Perturbation ", "",GC.H2.inv.mlt$Condition)

#CFU.inv.mlt$Time
```

### Plot H2  
```{r fig.width=6, fig.height=4}
ggplot(GC.H2.inv.mlt, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylab("% H2")
ggsave("01_wetlab/figs/h2_dynamics.pdf", h=4, w=8)

h2 <- ggplot(GC.H2.inv.mlt, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylab("% H2") +
  facet_grid(~study, scales = "free")
ggsave("01_wetlab/figs/h2_dynamics_study.pdf", h=4, w=8)
ggsave("01_wetlab/figs/h2_dynamics_study.png", h=4, w=8, dpi = 300)

```

###Combine plots of CO2 and H2
```{r}
Combineplot <- ggarrange(co2, h2,
                         labels = c("A", "B"),
                         ncol = 1, nrow = 2,
                         common.legend = TRUE, legend = "right")
print(Combineplot)
ggsave("01_wetlab/figs/combinec02h2.pdf", h=4, w=8)
ggsave("01_wetlab/figs/combinec02h2.png", h=4, w=8, dpi = 300)

```

###SFCA dynamics
```{r}

HPLC.AcetatemM.inv <- as.data.frame(read.csv("data/raw/wetlab/HPLCAcetatemMinv.csv", sep = ";"))
HPLC.LactatemM.inv <- as.data.frame(read.csv("data/raw/wetlab/HPLCLactatemMinv.csv", sep = ";"))

colnames(HPLC.AcetatemM.inv)[1] <- "AssayIdentifier"
colnames(HPLC.LactatemM.inv)[1] <- "AssayIdentifier"

scfa <- reshape2::melt(HPLC.AcetatemM.inv) %>% 
  mutate(SCFA = "Acetate") 
lct <- reshape2::melt(HPLC.LactatemM.inv) %>% 
  mutate(SCFA = "Lactate")

scfa <- bind_rows(scfa, lct) %>% 
  mutate(Time = as.numeric(gsub("t", "", variable))) %>% 
  select(-variable) %>% 
  left_join(meta_tabl)

scfa$Perturbation <-  gsub( "Perturbation ", "",scfa$Condition)

#CFU.inv.mlt$Time
```

### Plot SCFA  
```{r fig.width=10, fig.height=4}
ggplot(scfa, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(shape=21, size=3, alpha=0.5) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylab("mM") +
  facet_grid(~SCFA)

ggsave("01_wetlab/figs/scfa_dynamics.pdf", h=4, w=8)

ggplot(scfa, aes(x=Time, y=value, fill=Perturbation)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(aes(shape=SCFA, color=Perturbation)) +
  scale_fill_manual(values=mycols) +
  scale_x_continuous(breaks = c(0, 4, 8, 24)) +
  theme_biome_utils() +
  ylab("mM") +
  facet_grid(SCFA ~ study, scales = "free")

ggsave("01_wetlab/figs/scfa_dynamics_study.pdf", h=4, w=8)
ggsave("01_wetlab/figs/scfa_dynamics_study.png", h=4, w=8, dpi = 300)


```

#Phage transit in GIT simulation
## read metadata_GIT
```{r}
meta_tablgit <- as.data.frame(read.csv("data/raw/wetlab/metadatagit.csv", sep = ";"))
rownames(meta_tablgit) <- meta_tablgit$AssayIdentifier
```

## GIT Dynamics    
```{r}
git <- as.data.frame(read.csv("data/raw/wetlab/git.csv", sep = ";"))
colnames(git)[1] <- "AssayIdentifier"
colnames(git)[2] <- "Oral_2min"
colnames(git)[3] <- "Gastric_2h"
colnames(git)[4] <- "Intestinal_0h"
colnames(git)[5] <- "Intestinal_1h"
colnames(git)[6] <- "Intestinal_2h"

git.mlt <- reshape2::melt(git) %>%
  left_join(meta_tablgit)
git.mlt$study_f <- factor(git.mlt$study, levels = c("Control", "GIT")) 

```

### Plot PFU through the GIT transit
```{r fig.width=6, fig.height=4}
gitplot <- ggplot(git.mlt, aes(x=variable, y=value)) +
  geom_boxplot(aes(fill=as.factor(study_f))) +
  scale_fill_manual(values=mycolsgit) +
  theme_biome_utils() +
  xlab("Site") +
  ylab("Log10 PFU per mL") +
  ylim(0, 11.5) +
  labs(fill="Study")
gitplot
ggsave("01_wetlab/figs/git_dynamics_study.pdf", h=4, w=8)
ggsave("01_wetlab/figs/git_dynamics_study.png", h=4, w=8, dpi = 300)

ggplot(git.mlt, aes(x=variable, y=value, fill=study)) +
  stat_summary(geom="ribbon", alpha=0.6) +
  geom_point(aes(color=study)) +
  scale_fill_manual(values=mycolsgit) +
  theme_biome_utils() +
  ylab("Log10 PFU per mL") +
  xlab("Site") +
  ylim(0, 12)

```

### Statistic analysis
##PFU and CFU dynamics in in vitro fermentations
```{r}
library(rstatix)
library(formattable)
library(xlsx)

#PFU analysis
PFUstat0h <- PFU.inv %>%
  left_join(meta_tabl) %>%
  t_test( X0 ~ Condition , ref.group = "Perturbation Phage")
PFUstat0h

PFUstat4h <- PFU.inv %>%
  left_join(meta_tabl) %>%
  t_test( X4 ~ Condition , ref.group = "Perturbation Phage")
PFUstat4h

PFUstat8h <- PFU.inv %>%
  left_join(meta_tabl) %>%
  t_test( X8 ~ Condition , ref.group = "Perturbation Phage")
PFUstat8h

PFUstat24h <- PFU.inv %>%
  left_join(meta_tabl) %>%
  t_test( X24 ~ Condition , ref.group = "Perturbation Phage")
PFUstat24h

#CFU analysis
CFUstat0h <- CFU.inv %>%
  left_join(meta_tabl) %>%
  t_test( t0 ~ Condition , ref.group = "Perturbation E.coli")
CFUstat0h

CFUstat4h <- CFU.inv %>%
  left_join(meta_tabl) %>%
  t_test( t4 ~ Condition , ref.group = "Perturbation E.coli")
CFUstat4h

CFUstat8h <- CFU.inv %>%
  left_join(meta_tabl) %>%
  t_test( t8 ~ Condition , ref.group = "Perturbation E.coli")
CFUstat8h

CFUstat24h <- CFU.inv %>%
  left_join(meta_tabl) %>%
  t_test( t24 ~ Condition , ref.group = "Perturbation E.coli")
CFUstat24h

```

###making new plots with statistical analysis
```{r}
#Add * label for p.adj < 0.001
library(ggtext)
labelAst <- data.frame (Time= c(4, 8, 24),
                         value = c(7.5),lab = "*",
                        Perturbation = "Phage_E.coli")

newplot1 <- PFUplot + geom_text(data = labelAst,label = "*", size=7)
newplot1
ggsave("01_wetlab/figs/pfu_dynamics_stat.pdf", h=4, w=8)
ggsave("01_wetlab/figs/pfu_dynamics_stat.png", h=4, w=8, dpi = 300)

##Add * label for p.adj < 0.05
labelAstCFU <- data.frame (Time= c(8, 24),
                         value = c(7.5),lab = "*",
                        Perturbation = "Phage")

newplot2 <- CFUplot + geom_text(data = labelAstCFU,label = "*", size=7)
newplot2
ggsave("01_wetlab/figs/cfu_dynamics_stat.pdf", h=4, w=8)
ggsave("01_wetlab/figs/cfu_dynamics_stat.png", h=4, w=8, dpi = 300)
```

## Size and zeta analysis
```{r}
DLS <- as.data.frame(read.csv("data/raw/wetlab/sizeZeta.csv", sep = ";"))

sizestat <- DLS %>%
  t_test( size.nm ~ Condition , ref.group = "Control")
sizestat

zetastat <- DLS %>%
  t_test( zeta ~ Condition , ref.group = "Control")
zetastat

pdistat <- DLS %>%
  t_test( PDI ~ Condition , ref.group = "Control")
sizestat
```

#Statistical analyis for GIT plot
```{r}
#PFU analysis
oralstat <- git %>%
  left_join(meta_tablgit) %>%
  t_test( Oral_2min ~ study , ref.group = "Control")

oralstat

intestinal1stat <- git %>%
  left_join(meta_tablgit) %>%
  t_test( Intestinal_1h ~ study , ref.group = "Control")

intestinal1stat

intestinal2stat <- git %>%
  left_join(meta_tablgit) %>%
  t_test( Intestinal_2h ~ study , ref.group = "Control")

intestinal2stat

##Add * label for p.adj < 0.05
labelAstgit <- data.frame (study_f = (c("GIT")),
                           variable= c("Intestinal_1h", "Intestinal_2h"),
                         value = c(11),lab = "*")

newplot3 <- gitplot + geom_text(data = labelAstgit,label = "*", size=7, nudge_x = 0.2)
newplot3
ggsave("01_wetlab/figs/git_stat.pdf", h=4, w=8)

labelUDLgit <- data.frame (variable = "Gastric_2h",
                           value = c(0.4),lab = "UDL",
                           study=c("GIT"))

newplot4 <- newplot3 + geom_text(data = labelUDLgit,label = "UDL", size=4, nudge_x = 0.2)
newplot4

labelNAgit <- data.frame (variable = c("Intestinal_0h"),
                           value = c(0.4),lab = "N/A",
                           study_f=c("Control"))

newplot5 <- newplot4 + geom_text(data = labelUDLgit,label = "N/A", size=4, nudge_x = 0.8)
newplot5
ggsave("01_wetlab/figs/git_stat_info.pdf", h=4, w=8)
ggsave("01_wetlab/figs/git_stat_info.png", h=4, w=8, dpi = 300)
```

